---
Title: Harrison Manuscript - Richness and diversity code
Author: Georgia R Harrison
Date: March 17, 2021
output:
  html_document:
    theme: readable 
    toc: true
    toc_float: true
  pdf_document: default
---

# Analysis for Climbing impact manuscript
## Richness and Diversity Components
### Code by Georgia Harrison
#### March 11, 2021


### library packages
```{r, warning=FALSE, message=FALSE}
library(vegan) #community ecology
library(ggplot2) #data viz
library(dplyr) #data manipulation
library(tidyr) #data manipulation
library(ggpubr) #for combining multiple ggplots into one figure
library(ddpcr) #for joining data sets
```


## Import the data 
Remove bluff mountain, only want face plots
```{R}
data <- read.csv("~/Cliffs/Manuscript/HARRISON_LGWA_MATRIX.csv", header=TRUE)

data <- data %>%
  group_by(T_LOCATION) %>% 
  filter(T_LOCATION %in% c("HB", "TR")| n() == 1)
#remove bluff mtn

data <- data %>%
  group_by(P_LOCATION) %>% 
  filter(P_LOCATION %in% c("FACE")| n() == 1)
```


import data in long format too 
```{R}
long <- read.csv("~/Cliffs/Manuscript/Long_data.csv", header=TRUE)
```



## Create matrix for each taxa group 
First, determine which columns start and end each taxa group and then create a matrix for each taxa group
```{r, collapse=TRUE}
which( colnames(data)=="BLAFLA") #54
which( colnames(data)=="YELPDR") #133
which( colnames(data)=="Moss_2") #134
which( colnames(data)=="Wei_controversa") #174
which( colnames(data)=="Ace_saccharum") #175
which( colnames(data)=="Plant_5") #212


total.abundance.matrix <- data[,54:212]
lichen.abundance.matrix <- data[,54:133]
moss.abundance.matrix <- data [,134:174]
plant.abundance.matrix <- data [,175:212]

```



### Create a new data frame to input richness, diveristy and abundance values (but keep the important metadata)
```{r}
indic <- data[,c("NAME","ROUTE","PITCH_NUM", "CL_UNCL", "PLOT_NUM", "P_LOCATION", "T_LOCATION", "ROUTE_TYPE", 
                   "GRADE_CAT", "STARS", "TOT_HEIGHT", "SLOPE", "ASPECT_NORTH", "ASPECT_EAST")]
indic$T_LOCATION <- as.factor((indic$T_LOCATION))
indic$CL_UNCL <- as.factor((indic$CL_UNCL))
```


## Determine speciess richnes, shannon and total abundance for each 
the cover will be used as a proxy for abundance 
```{r}
#all
indic$AllRichness <- rowSums(total.abundance.matrix>0)
indic$AllShannon <- diversity(total.abundance.matrix) # shannon is default
indic$AllCover <- rowSums(total.abundance.matrix) 

#lichens
indic$LRichness <- rowSums(lichen.abundance.matrix>0)
indic$LShannon <- diversity(lichen.abundance.matrix) # shannon is default
indic$LCover <- rowSums(lichen.abundance.matrix)

#mosses
indic$MRichness <- rowSums(moss.abundance.matrix>0)
indic$MShannon <- diversity(moss.abundance.matrix) # shannon is default
indic$MCover <- rowSums(moss.abundance.matrix)

#plants
indic$PRichness <- rowSums(plant.abundance.matrix>0)
indic$PShannon <- diversity(plant.abundance.matrix) # shannon is default
indic$PCover <- rowSums(plant.abundance.matrix)
```



# Compare richness and diversity by climbing and site
Using ANOVAs
```{r}
##all species

Ranova <- aov(AllRichness~T_LOCATION*CL_UNCL, data=indic)
summary(Ranova)
TukeyHSD(Ranova)

Danova <- aov(AllShannon~T_LOCATION*CL_UNCL, data=indic)
summary(Danova)
TukeyHSD(Danova)

Canova <- aov(AllCover~T_LOCATION*CL_UNCL, data=indic)
summary(Canova)
TukeyHSD(Canova)


#lichens only
Ranova <- aov(LRichness~T_LOCATION*CL_UNCL, data=indic)
summary(Ranova)
TukeyHSD(Ranova)

Danova <- aov(LShannon~T_LOCATION*CL_UNCL, data=indic)
summary(Danova)
TukeyHSD(Danova)

Canova <- aov(LCover~T_LOCATION*CL_UNCL, data=indic)
summary(Canova)
TukeyHSD(Canova)


###PLANTS
Ranova <- aov(PRichness~T_LOCATION*CL_UNCL, data=indic)
summary(Ranova)
TukeyHSD(Ranova)

Danova <- aov(PShannon~T_LOCATION*CL_UNCL, data=indic)
summary(Danova)
TukeyHSD(Danova)
Canova <- aov(PCover~T_LOCATION*CL_UNCL, data=indic)
summary(Canova)
TukeyHSD(Canova)


##MOSSES
Ranova <- aov(MRichness~T_LOCATION*CL_UNCL, data=indic)
summary(Ranova)
TukeyHSD(Ranova)

Danova <- aov(MShannon~T_LOCATION*CL_UNCL, data=indic)
summary(Danova)
TukeyHSD(Danova)
Canova <- aov(MCover~T_LOCATION*CL_UNCL, data=indic)
summary(Canova)
TukeyHSD(Canova)
```




# Figures for diversity, richness and cover for all taxa by climbing and site
```{r}
theme_set(
    theme(
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.background = element_blank(), 
      axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), 
      axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), 
      legend.text = element_text(size=16, colour= "black")
      ))

#richness
p <- ggplot(indic, aes(x=T_LOCATION, y=AllRichness, fill=CL_UNCL))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Species Richness", x="Site")
pr <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))

#diversity
p <- ggplot(indic, aes(x=T_LOCATION, y=AllShannon, fill=CL_UNCL))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Shannon Diversity", x="Site")
pd <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))

#abundance
p <- ggplot(indic, aes(x=T_LOCATION, y=AllCover, fill=CL_UNCL))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Relative Abundance", x="Site")
pa <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))


all_taxa_summary <- ggarrange(pr, pd, pa, ncol=3, common.legend = TRUE)
all_taxa_summary 

ggsave("alltaxa_summary.jpeg", plot = all_taxa_summary,  device = "jpeg", width = 8, height = 4, dpi = 300, scale = 1)

```




## Figure for species richness for all groups 
If needed, we could do the same thing for diversity and cover - but might be better to just put this info into a table?
```{r}
theme_set(
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
      panel.background = element_blank(), axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), legend.text = element_text(size=16, colour= "black")))


all <- ggplot(indic, aes(x=T_LOCATION, y=AllRichness, fill=factor(CL_UNCL)))
all <- all + geom_violin(trim=TRUE)+theme_classic()
all <- all + labs(y="Species Richness", x="Site", title="All Taxa")
all <- all + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(all)

L <- ggplot(indic, aes(x=T_LOCATION, y=LRichness, fill=factor(CL_UNCL)))
L <- L + geom_violin(trim=TRUE) +theme_classic()
L <- L + labs(y="Species Richness", x="Site", title="Lichens")
L <- L + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(L)

M <- ggplot(indic, aes(x=T_LOCATION, y=MRichness, fill=factor(CL_UNCL)))
M <- M + geom_violin(trim=TRUE)+theme_classic()
M <- M + labs(y="Species Richness", x="Site", title="Bryophytes")
M <- M + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(M)

P <- ggplot(indic, aes(x=T_LOCATION, y=PRichness, fill=factor(CL_UNCL)))
P <- P + geom_violin(trim=TRUE)+theme_classic()
P <- P + labs(y="Species Richness", x="Site", title="Vascular Plants")
P <- P + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(P)


#combine into one figure
combined_rich <- ggarrange(all, L, M, P, common.legend = TRUE)
combined_rich 
ggsave("combined_rich.jpeg", plot = combined_rich,  device = "jpeg", width = 8, height = 6, dpi = 300, scale = 1)
```




## Figure for diversity for all groups 
If needed, we could do the same thing for diversity and cover - but might be better to just put this info into a table?
```{r}
theme_set(
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
      panel.background = element_blank(), axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), legend.text = element_text(size=16, colour= "black")))


all <- ggplot(indic, aes(x=T_LOCATION, y=AllShannon, fill=factor(CL_UNCL)))
all <- all + geom_violin(trim=TRUE)+theme_classic()
all <- all + labs(y="Species Diversity", x="Site", title="All Taxa")
all <- all + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(all)

L <- ggplot(indic, aes(x=T_LOCATION, y=LShannon, fill=factor(CL_UNCL)))
L <- L + geom_violin(trim=TRUE) +theme_classic()
L <- L + labs(y="Species Diversity", x="Site", title="Lichens")
L <- L + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(L)

M <- ggplot(indic, aes(x=T_LOCATION, y=MShannon, fill=factor(CL_UNCL)))
M <- M + geom_violin(trim=TRUE)+theme_classic()
M <- M + labs(y="Species Diversity", x="Site", title="Bryophytes")
M <- M + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(M)


P <- ggplot(indic, aes(x=T_LOCATION, y=PShannon, fill=factor(CL_UNCL)))
P <- P + geom_violin(trim=TRUE)+theme_classic()
P <- P + labs(y="Species Diversity", x="Site", title="Vascular Plants")
P <- P + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(P)


#combine into one figure
combined_div <- ggarrange(all, L, M, P, common.legend = TRUE)
combined_div
ggsave("combined_div.jpeg", plot = combined_div,  device = "jpeg", width = 8, height = 6, dpi = 300, scale = 1)
```



## Figure for cover for all groups 
If needed, we could do the same thing for diversity and cover - but might be better to just put this info into a table?
```{r}
theme_set(
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
      panel.background = element_blank(), axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), legend.text = element_text(size=16, colour= "black")))


all <- ggplot(indic, aes(x=T_LOCATION, y=AllCover, fill=factor(CL_UNCL)))
all <- all + geom_violin(trim=TRUE)+theme_classic()
all <- all + labs(y="Relative abundance", x="Site", title="All Taxa")
all <- all + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(all)

L <- ggplot(indic, aes(x=T_LOCATION, y=LCover, fill=factor(CL_UNCL)))
L <- L + geom_violin(trim=TRUE) +theme_classic()
L <- L + labs(y="Relative abundance", x="Site", title="Lichens")
L <- L + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(L)

M <- ggplot(indic, aes(x=T_LOCATION, y=MCover, fill=factor(CL_UNCL)))
M <- M + geom_violin(trim=TRUE)+theme_classic()
M <- M + labs(y="Relative abundance", x="Site", title="Bryophytes")
M <- M + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(M)

P <- ggplot(indic, aes(x=T_LOCATION, y=PCover, fill=factor(CL_UNCL)))
P <- P + geom_violin(trim=TRUE)+theme_classic()
P <- P + labs(y="Relative abundance", x="Site", title="Vascular Plants")
P <- P + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(P)


#combine into one figure
combined_cover <- ggarrange(all, L, M, P, common.legend = TRUE)
combined_cover 
ggsave("combined_cover.jpeg", plot = combined_cover,  device = "jpeg", width = 8, height = 6, dpi = 300, scale = 1)
```





# Difference by climbing grade
```{r}
anova <- aov(AllShannon~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

anova <- aov(AllRichness~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

anova <- aov(AllCover~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)
```
Taxa specific differences by climbing grade
```{r}
#LIchens only
anova <- aov(LShannon~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

anova <- aov(LRichness~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

 #MOSSES only
anova <- aov(MShannon~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

anova <- aov(MRichness~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

#plants only
anova <- aov(PShannon~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)

anova <- aov(PRichness~GRADE_CAT, data = indic)
summary(anova)
TukeyHSD(anova)
```


# figures for by climbing grade category
### All taxa richness, diversity and cover by climbing grade
```{r}
level_order <- c("EASY", "MODERATE", "ADVANCED", "UNCLIMBED")
#how to arranged factors of x axis in order of increasing difficulty


d<- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=AllShannon, fill=CL_UNCL)) +
      labs(y="Shannon Diversity", x="Climbing Grade") +
      geom_violin(trim=TRUE) +
      scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))+
      theme_classic()+
      theme(axis.text.x = element_text(angle=45, hjust=1))+
      ylim(0,3)

r<- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=AllRichness, fill=CL_UNCL)) +
      labs(y="Species Richness", x="Climbing Grade") +
      geom_violin(trim=TRUE) +
      scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))+
      theme_classic()+
      theme(axis.text.x = element_text(angle=45, hjust=1))+
      ylim(0,18)

a <- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=AllCover, fill=CL_UNCL)) +
      labs(y="Relative Abundance", x="Climbing Grade") +
      geom_violin(trim=TRUE) +
      scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))+
      theme_classic()+
      theme(axis.text.x = element_text(angle=45, hjust=1))

climbing_level <- ggarrange(r, d, a, ncol=3,common.legend = TRUE)
climbing_level

ggsave("climbing_grade.jpeg", plot = climbing_level,  device = "jpeg", width = 8, height = 4, dpi = 300, scale = 1)
```


### Richness of all taxa groups by climbing grade
```{r}
theme_set(
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
      panel.background = element_blank(), axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), legend.text = element_text(size=16, colour= "black")))


all <- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=AllRichness, fill=CL_UNCL))
all <- all + geom_violin(trim=FALSE)+theme_classic()
all <- all + labs(y="Species Richness", x="Site", title="All Taxa")
all <- all + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(all)

L <- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=LRichness, fill=CL_UNCL))
L <- L + geom_violin(trim=FALSE) +theme_classic()
L <- L + labs(y="Species Richness", x="Site", title="Lichens")
L <- L + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(L)

M <- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=MRichness, fill=CL_UNCL))
M <- M + geom_violin(trim=FALSE)+theme_classic()
M <- M + labs(y="Species Richness", x="Site", title="Bryophytes")
M <- M + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(M)

P <- ggplot(indic, aes(x= factor(GRADE_CAT, level = level_order), y=PRichness, fill=CL_UNCL))
P <- P + geom_violin(trim=FALSE)+theme_classic()
P <- P + labs(y="Species Richness", x="Site", title="Vascular Plants")
P <- P + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(P)


#combine into one figure
combined_rich_grade <- ggarrange(all, L, M, P, common.legend = TRUE)
combined_rich_grade 
```



## create a summary table for richness, cover, and diveristy 
```{r}
indic$T_LOCATION <- as.factor((indic$T_LOCATION))
indic$CL_UNCL <- as.factor((indic$CL_UNCL))
t_all <- indic %>%
  group_by(T_LOCATION, CL_UNCL) %>%
    summarise(
          mean_rich=mean(AllRichness), sd_rich=sd(AllRichness), min_rich=min(AllRichness), max_rich=max(AllRichness),
          mean_div=mean(AllShannon), sd_div=sd(AllShannon), min_div=min(AllShannon), max_div=max(AllShannon),
          mean_abun=mean(AllCover), sd_abun=sd(AllCover), min_abun=min(AllCover), max_abun=max(AllCover)
          )
head(t_all)
```


```{r}
t_lichen <- indic %>%
  group_by(T_LOCATION, CL_UNCL) %>%
    summarise(
          mean_rich=mean(LRichness), sd_rich=sd(LRichness), min_rich=min(LRichness), max_rich=max(LRichness),
          mean_div=mean(LShannon), sd_div=sd(LShannon), min_div=min(LShannon), max_div=max(LShannon),
          mean_abun=mean(LCover), sd_abun=sd(LCover), min_abun=min(LCover), max_abun=max(LCover)
          )
t_lichen$taxa <- "lichen"

t_moss <- indic %>%
  group_by(T_LOCATION, CL_UNCL) %>%
    summarise(
          mean_rich=mean(MRichness), sd_rich=sd(MRichness), min_rich=min(MRichness), max_rich=max(MRichness),
          mean_div=mean(MShannon), sd_div=sd(MShannon), min_div=min(MShannon), max_div=max(MShannon),
          mean_abun=mean(MCover), sd_abun=sd(MCover), min_abun=min(MCover), max_abun=max(MCover)
          )
t_moss$taxa <- "moss"

t_plants <- indic %>%
  group_by(T_LOCATION, CL_UNCL) %>%
    summarise(
          mean_rich=mean(PRichness), sd_rich=sd(PRichness), min_rich=min(PRichness), max_rich=max(PRichness),
          mean_div=mean(PShannon), sd_div=sd(PShannon), min_div=min(PShannon), max_div=max(PShannon),
          mean_abun=mean(PCover), sd_abun=sd(PCover), min_abun=min(PCover), max_abun=max(PCover)
          )
t_plants$taxa <- "plants"
```


```{r}
summary_stats<- 
  merge_dfs_overwrite_col(
    t_all, t_lichen, 
    bycol = c("CL_UNCL", "T_LOCATION"))
head(summary_stats)
summary_stats<- 
  merge_dfs_overwrite_col(
    summary_stats, t_moss, 
    bycol = c("CL_UNCL", "T_LOCATION", 
              "mean_rich", "sd_rich", "min_rich", "max_rich", 
              "mean_div", "sd_div", "min_div", "max_div", 
              "mean_abun", "sd_abun", "min_abun", "max_abun"))
summary_stats<- 
  merge_dfs_overwrite_col(
    summary_stats, t_plants, 
    bycol = c("CL_UNCL", "T_LOCATION"))

```








## Rank abundance curves 
```{r}
library(BiodiversityR)
library(tibble)


#load in community data and environment
data1 <- data %>% column_to_rownames(var="NAME")
com <- data1[,53:211]
head(com)

#environmental factors: climbing and site
env <- data1[,c("CL_UNCL", "T_LOCATION")]
env$Climbing <- as.factor(env$CL_UNCL)
env$Site <- as.factor(env$T_LOCATION)

RankAbun1 <- rankabundance(com, y=env)
RankAbun1
rankabunplot(RankAbun1, scale='abundance', addit=FALSE, specnames=c(1,2,3))
```


Rank abundance curves by climbing and site
```{r}

rankabuncomp(com, y=env, factor="Site", scale='proportion', legend=FALSE)
RA.Site <- rankabuncomp(com, y=env, factor='Site', 
    return.data=TRUE, specnames=c(1:10), legend=FALSE)
RA.Site


rankabuncomp(com, y=env, factor="Climbing", scale='proportion', legend=FALSE)
RA.Cl <- rankabuncomp(com, y=env, factor='Climbing', 
    return.data=TRUE, specnames=c(1:10), legend=FALSE)
RA.Cl

```

FIgures for rank abundance curves by site and climbing 
```{r}
library(ggrepel)

plotgg2 <- ggplot(data=RA.Site, aes(x = rank, y = abundance)) 
plotgg2 <- plotgg2 + scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) +
          scale_y_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) +
          geom_line(aes(colour=Grouping), size=1) +
          geom_point(aes(colour=Grouping), size=5, alpha=0.7) +
          geom_text_repel(data=subset(RA.Site, labelit == TRUE),
          aes(label=species),
          angle=45, nudge_x=1, nudge_y=1, show.legend=FALSE) +
          scale_color_brewer(palette = "Dark2") +
          labs(x = "rank", y = "abundance", colour = "Site")
plotgg2



plotgg3 <- ggplot(data=RA.Cl, aes(x = rank, y = abundance)) 
plotgg3 <- plotgg3 + scale_x_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) +
          scale_y_continuous(expand=c(0, 1), sec.axis = dup_axis(labels=NULL, name=NULL)) +
          geom_line(aes(colour=Grouping), size=1) +
          geom_point(aes(colour=Grouping), size=5, alpha=0.7) +
          geom_text_repel(data=subset(RA.Cl, labelit == TRUE),
          aes(label=species),
          angle=45, nudge_x=1, nudge_y=1, show.legend=FALSE) +
          scale_color_brewer(palette = "Dark2") +
          labs(x = "rank", y = "abundance", colour = "Site")
plotgg3
```

