---
Title: Harrison Manuscript - Lichen functional groups
Author: Georgia R Harrison
Date: March 17, 2021
output:
  html_document:
    theme: readable 
    toc: true
    toc_float: true
  pdf_document: default
---

# Analysis for Climbing impact manuscript
## Lichen functional groups
### Code by Georgia Harrison
#### March 11, 2021


### library packages
```{r, warning=FALSE, message=FALSE}
# install.packages("vegan")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("tidyr")
# install.packages("ggpubr")
# install.packages("ddpcr")
# install.packages("ellipsis")

library(vegan) #community ecology
library(ggplot2) #data viz
library(dplyr) #data manipulation
library(tidyr) #data manipulation
library(ggpubr) #for combining multiple ggplots into one figure
library(ddpcr) #for joining data sets
```


## lichen functional groups
bring in the data, make functional groupings a factor. Using long data set because that was the long way to add in functional groups.
```{r}
long <- read.csv("~/Cliffs/Manuscript/Long_data.csv", header=TRUE)

long <- long %>%
  group_by(T_LOCATION) %>% 
  filter(T_LOCATION %in% c("HB", "TR")| n() == 1)
#remove bluff mtn

long <- long %>%
  group_by(P_LOCATION) %>% 
  filter(P_LOCATION %in% c("FACE")| n() == 1)
```
### we only want the lichens
```{r}
lichen <- long[long$TAXA=="L", ]

lichen$Functional <- factor(lichen$Functional)
summary(lichen$Functional)

View(lichen)

which(colnames(lichen)=="DATE") #2
which(colnames(lichen)=="Abundance") #18
lichen <- lichen[,2:18]

```
create a list of columns which we care about (environmetnal factors)
```{r}
env <- c("CL_UNCL", "T_LOCATION", "SLOPE", "ASPECT_NORTH", "ASPECT_EAST", "ROUTE","PLOT_NUM")

```


## Create richness, diversity and cover values for each fx group
```{r, collapse = TRUE}
crust <- lichen[lichen$Functional=="CRUST",]
crust <- spread(crust, Species, Abundance)

which(colnames(crust)=="BLBRDOT") #16
which(colnames(crust)=="YELLWART") #44
crust.matrix <- crust[,16:44]

crust_incic <- crust [,env]

crust_incic$crust.rich <- rowSums(crust.matrix>0)
crust_incic$crust.div <- diversity(crust.matrix) # shannon is default
crust_incic$crust.cov <- rowSums(crust.matrix) 



fruit <- lichen[lichen$Functional=="FRUIT",]
fruit <- spread(fruit, Species, Abundance)
which(colnames(fruit)=="WHITREIN") #31
fruit.matrix <- fruit[,16:31]

fruit_indic <- fruit [,env]

fruit_indic$fruit.rich <- rowSums(fruit.matrix>0)
fruit_indic$fruit.div <- diversity(fruit.matrix) # shannon is default
fruit_indic$fruit.cov <- rowSums(fruit.matrix) 



lobe <- lichen[lichen$Functional=="LOBE",]
lobe <- spread(lobe, Species, Abundance)
which(colnames(lobe)=="WHITEFOL") #31
lobe.matrix <- lobe[,16:36]

lobe_indic <- lobe [,env]

lobe_indic$lobe.rich <- rowSums(lobe.matrix>0)
lobe_indic$lobe.div <- diversity(lobe.matrix) # shannon is default
lobe_indic$lobe.cov <- rowSums(lobe.matrix) 



powder <- lichen[lichen$Functional=="POWDER",]
powder <- spread(powder, Species, Abundance)
which(colnames(powder)=="YELPDR") #25
powder.matrix <- powder[,16:25]

powder_indic <- powder [,env]

powder_indic$powder.rich <- rowSums(powder.matrix>0)
powder_indic$powder.div <- diversity(powder.matrix) # shannon is default
powder_indic$powder.cov <- rowSums(powder.matrix) 



umb <- lichen[lichen$Functional=="UMB",]
umb <- spread(umb, Species, Abundance)
which(colnames(umb)=="TOADSK") #19
umb.matrix <- umb[,16:19]


umb_indic <- umb [,env]

umb_indic$umb.rich <- rowSums(umb.matrix>0)
umb_indic$umb.div <- diversity(umb.matrix) # shannon is default
umb_indic$umb.cov <- rowSums(umb.matrix) 
```


merge all the lichen functional groups to one data set
```{r}
L_merge <- merge(crust_incic, fruit_indic, by.x=env, by.y=env)
L_merge <- merge(L_merge, lobe_indic, by.x=env, by.y=env)
L_merge <- merge(L_merge, powder_indic, by.x=env, by.y=env)
L_merge <- merge(L_merge, umb_indic, by.x=env, by.y=env)

head(L_merge)

```



## tidy up the data for analysis and figures
1. Subset out richness, diversity and cover for each fx group (find column numbers). Still include plot metadata
2. rename functional groups to consistent names
3. convert to long format
4. Merge all together so you have a long dataset with richness, diversity and cover for each functional group at each plot
    + note that we still have individual richness, diversity and cover data frames if we need to run analyisis in those formats
```{r, collapse = TRUE}
#use head(L_merge) to keep track of column labels

#metadata
which(colnames(L_merge)=="CL_UNCL") #1
which(colnames(L_merge)=="PLOT_NUM") #7

listofnames <- c("crustose", "fruticose", "lobed foliose", "powder", "umbilicate foliose")

#richness
which(colnames(L_merge)=="crust.rich") #8
which(colnames(L_merge)=="umb.rich") #20
#r <- c(1:7, 8, 11, 14, 17, 20) #these columms with metadata and richness
richness <- L_merge[, c(1:7, 8, 11, 14, 17, 20)]
colnames(richness)[8:12] <- listofnames #rename columns

richness <- gather(richness, fx, richness, 8:12) #long format
#fx is functional group


#diversity
which(colnames(L_merge)=="crust.div") #9
which(colnames(L_merge)=="umb.div") #21
#d<- c(1:7, 9, 12, 15, 18, 21)
diversity <- L_merge[, c(1:7, 9, 12, 15, 18, 21)]
colnames(diversity)[8:12] <- listofnames #rename columns

diversity <- gather(diversity, fx, diversity, 8:12) #long format
  
#cover
which(colnames(L_merge)=="crust.cov") #10
which(colnames(L_merge)=="umb.cov") #22
#c<- c(1:7, 10, 13, 16, 19, 22)
cover <- L_merge[,c(1:7, 10, 13, 16, 19, 22)]
colnames(cover)[8:12] <- listofnames #rename columns

cover <- gather(cover, fx, cover, 8:12) #long format

  


#combine them all together
env1 <- c("CL_UNCL", "T_LOCATION", "SLOPE", "ASPECT_NORTH", "ASPECT_EAST", "ROUTE","PLOT_NUM", "fx")
lindic <- merge(richness, diversity, by.x=env1, by.y=env1)
lindic <- merge(lindic, cover, by.x=env1, by.y=env1)
head(lindic)
```


## Figures for lichen fx groups
```{r}
theme_set(
    theme(
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.background = element_blank(), 
      axis.text = element_text(size=11, colour = "black"),  
      axis.line = element_line(colour = "black"), 
      axis.title = element_text(size=16, colour = "black"),
      legend.title = element_text(size = 16, colour = "black"), 
      legend.text = element_text(size=16, colour= "black"),
      strip.text.x = element_text(size=16)
      ))

#richness
p <- ggplot(lindic, aes(x=T_LOCATION, y=richness, fill=CL_UNCL))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Species Richness", x="Site", title="Species richness by lichen functional group")
p <- p + facet_wrap(~fx)
pr <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(pr)

#diversity
p <- ggplot(lindic, aes(x=T_LOCATION, y=diversity, fill=CL_UNCL))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Shannon Diversity Index", x="Site", title="Diversity by lichen functional group")
p <- p + facet_wrap(~fx)
pd <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(pd)

#abundance
p <- ggplot(lindic, aes(x=T_LOCATION, y=cover, fill=CL_UNCL))
p <- p + geom_violin(trim=TRUE)+theme_classic()
p <- p + labs(y="Relative abundance", x="Site")
p <- p + theme(legend.position = "top")
p <- p + facet_wrap(~fx)
p <- p + ylim(0,35)
pc <- p + scale_fill_grey(name="Climbing", labels = c("Climbed", "Unclimbed"))
plot(pc)



ggsave("lichen_funct_cover.jpeg", plot = pc,  
       path = "~/Cliffs/Manuscript/Figure export from R",
       device = "jpeg", width = 8, height = 5, dpi = 300, scale = 1)
```

**might also be interesting to flip climbing and site**
```{r}
p <- ggplot(lindic, aes(x=CL_UNCL, y=richness, fill=T_LOCATION))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Species Richness", x="Site", title="Species richness by lichen functional group")
p <- p + facet_wrap(~fx)
pr <- p + scale_fill_grey(name="Site", labels = c("HB", "TR"))
plot(pr)

p <- ggplot(lindic, aes(x=CL_UNCL, y=cover, fill=T_LOCATION))
p <- p + geom_violin(trim=FALSE)+theme_classic()
p <- p + labs(y="Relative abundance", x="Site", title="Cover by lichen functional group")
p <- p + facet_wrap(~fx)
pc <- p + scale_fill_grey(name="Site", labels = c("HB", "TR"))
plot(pc)


```

## anovas on lichen functional groups 
#### richness cover and diversity by functional group
```{r}
aov <-aov(richness~fx, data=lindic)
summary(aov)
TukeyHSD(aov)
aov <-aov(diversity~fx, data=lindic)
summary(aov)
TukeyHSD(aov)
aov <-aov(cover~fx, data=lindic)
summary(aov)
TukeyHSD(aov)
```

### by climbing and site
```{r}
aov <-aov(richness~CL_UNCL*T_LOCATION, data=lindic)
summary(aov)
TukeyHSD(aov)
aov <-aov(diversity~CL_UNCL*T_LOCATION, data=lindic)
summary(aov)
TukeyHSD(aov)
aov <-aov(cover~CL_UNCL*T_LOCATION, data=lindic)
summary(aov)
TukeyHSD(aov)
```


### Each functioanl group 
Need to use L_merge df for this.  
Just **crust** first
```{r}
aov <-aov(crust.rich~CL_UNCL*T_LOCATION, data=L_merge)
summary(aov)
TukeyHSD(aov)
aov <-aov(crust.div~CL_UNCL*T_LOCATION, data=L_merge)
summary(aov)
TukeyHSD(aov)
aov <-aov(crust.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(aov)
TukeyHSD(aov)
```

### Cover of each functional group by site and climbing

```{r}
caov <-aov(crust.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(caov)
TukeyHSD(caov)
caov <-aov(powder.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(caov)
TukeyHSD(caov)
caov <-aov(fruit.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(caov)
TukeyHSD(caov)
caov <-aov(lobe.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(caov)
TukeyHSD(caov)
caov <-aov(umb.cov~CL_UNCL*T_LOCATION, data=L_merge)
summary(caov)
TukeyHSD(caov)
```
